---
title: "Trifecta of CRISPR, RNAi and drug screen analysis"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(ppcor)

#PATHS
BASE=dirname(rstudioapi::getSourceEditorContext()$path)

#function for correlating drug response data
corDrugResponse<-function(hit_list,tau_thresh)
{
response<-c()
for(i in 1:nrow(hit_list))
{
  gname<-strsplit(hit_list$gene[i],"\\.")[[1]][1]
  drugs<-drug_lookup[drug_lookup$target==gname,"drug"]
  sig<-hit_list$signature[i]
  for(d in drugs)
  {
    pdat<-left_join(drug_screen[drug_screen$name==d,],
                    data.frame(depmap_id=rownames(cell_exp_filt_drug),cell_exp_filt_drug),
                    by="depmap_id")
    pdat<-left_join(pdat,cell_line_info,by="depmap_id")
    pdat<-pdat[pdat$passed_str_profiling.y==T,]
    pdat<-pdat[!is.na(pdat[,paste0("CX",sig)]),]
    res<-cor.test(pdat[,"auc"],pdat[,paste0("CX",sig)],method="k",alternative = "less")
    if(res$estimate< tau_thresh)
    {
      response<-rbind(response,c(sig,gname,d,res$estimate,res$p.value))
    }
  }
}

colnames(response)<-c("signature","target","drug","tau","p.value")
response<-cbind(response,q.value=p.adjust(response[,5],method="BH"))
response<-data.frame(response,stringsAsFactors = F)
response<-response[order(response$signature,response$q.value),]
response
}
```

##Load cell line signature exposures
```{r}
#load signature exposures
cell_exp<-readRDS(paste0(BASE,"/data/4_TCGA_PCAWG_Exposures_to_TCGA_Signatures_NewASCAT_NewCXnames_cellLines_penalty70.rds"))

#shrink low exposures to zero
thresh<-read.table(paste0(BASE,"/data/3_Boxplots_per_sig_fullTCGA_1000sims_10pGaussian_10pSamplePoisson_NewCXnames.txt"),sep="\t",header=T)
cell_exp_binary<-t(apply(cell_exp,1,function(x){x<thresh[,4]}))
cell_exp[cell_exp_binary]<-0

#load identifier mapping file
cell_id_mapping<-read.table(paste0(BASE,"/data/cellLine_CEL_file_mapping.tsv"),sep="\t",header=T,stringsAsFactors=F)
rownames(cell_exp)<-cell_id_mapping[match(rownames(cell_exp),cell_id_mapping[cell_id_mapping$study=="CCLE","fileid"]),"cellid"]

#remove duplicate entries
#cell_exp<-cell_exp[rev(rownames(cell_exp)),]
cell_exp<-cell_exp[!duplicated(rownames(cell_exp)),]

samp_info<-read.csv(paste0(BASE,"/data/sample_info.csv"),header=T)
rownames(cell_exp)<-samp_info[unlist(sapply(rownames(cell_exp),function(x){grep(x,samp_info$cell_line_name,ignore.case = T)[1]})),1]
cell_exp<-cell_exp[!is.na(rownames(cell_exp)),]
cell_exp<-cell_exp[!duplicated(rownames(cell_exp)),]

print("Total number of cell lines with signature activities:")
nrow(cell_exp)
```
##Load high-throughput drug screen data
```{r}
#load drug screen data & sample info --> download from https://depmap.org/repurposing/
drug_screen<-read.csv(paste0(BASE,"/data/secondary-screen-dose-response-curve-parameters.csv"),stringsAsFactors = F)
drug_screen<-drug_screen[drug_screen$screen_id=="HTS002",]
cell_line_info<-read.csv(paste0(BASE,"/data/secondary-screen-cell-line-info.csv"))
cell_exp_filt_drug<-cell_exp[rownames(cell_exp)%in%drug_screen$depmap_id,]

drug_lookup<-c()
unique_target<-unique(drug_screen[,c("name","target")])
for(i in 1:nrow(unique_target))
{
  drug<-unique_target[i,"name"]
  targets<-unlist(strsplit(unique_target[i,"target"],", "))
  drug_lookup<-rbind(drug_lookup,cbind(drug,targets))
}

drug_lookup<-unique(drug_lookup)
drug_lookup<-data.frame(drug_lookup)
colnames(drug_lookup)<-c("drug","target")
drug_lookup<-drug_lookup[!is.na(drug_lookup$target),]

print("Number of cell lines with drug response data:")
nrow(cell_exp_filt_drug)
```

##Load CRISPR screen data
```{r}
#load crispr knockout data --> download from https://depmap.org/portal/download/
gene_effect<-read.csv(paste0(BASE,"/data/CRISPR_gene_effect.csv"),row.names = 1)

#filter to matching cell line data between KO and exposures
cell_exp_filt_KO<-cell_exp[rownames(cell_exp)%in%rownames(gene_effect),]
cell_exp_binary_filt_KO<-cell_exp_binary[rownames(cell_exp_binary)%in%rownames(gene_effect),]
gene_effect_filt<-gene_effect[match(rownames(cell_exp_filt_KO),rownames(gene_effect)),]
print("Total number of cell lines with signatures and CRISPR screen data:")
nrow(gene_effect_filt)
```

##Load RNAi screen data
```{r}
#load rnai knockdown data --> download from https://depmap.org/portal/download/
gene_knockdown<-read.csv(paste0(BASE,"/data/D2_combined_gene_dep_scores.csv"),row.names = 1,header=F,stringsAsFactors = F)
knockdown_samp_info<-read.csv(paste0(BASE,"/data/DEMETER_sample_info.csv"))
colnames(gene_knockdown)<-samp_info[match(gene_knockdown[1,],samp_info[,4]),1]
gene_knockdown<-t(gene_knockdown[-1,])
gene_knockdown<-data.frame(gene_knockdown[!is.na(rownames(gene_knockdown)),],stringsAsFactors = F)

#filter to matching cell line data between KO and exposures
cell_exp_filt_KD<-cell_exp[rownames(cell_exp)%in%rownames(gene_knockdown),]
gene_knockdown_filt<-gene_knockdown[match(rownames(cell_exp_filt_KD),rownames(gene_knockdown)),]
print("Total number of cell lines with signatures and RNAi screen data:")
nrow(gene_effect_filt)
```

#Analysis
##Correlation between CRISPR gene essentiality score and signature activity
```{r}
results_cor<-c()
for(j in 1:ncol(cell_exp_filt_KO))
{
  res_sig<-c()
  for(i in 1:ncol(gene_effect_filt))
  {
    res<-cor.test(gene_effect_filt[,i],cell_exp_filt_KO[,j],method="k",alternative = "less")
      if(res$estimate< -0.07)
      {
        res_sig<-rbind(res_sig,c(j,colnames(gene_effect_filt)[i],
                               res$estimate,res$p.value))
      }
  }
  res_sig<-data.frame(res_sig,stringsAsFactors = F)
  res_sig<-cbind(res_sig,q.value=p.adjust(res_sig[,4],method="BH"))
  results_cor<-rbind(results_cor,res_sig)
}

colnames(results_cor)<-c("signature","gene","rho","p.value","q.value")
results_cor<-data.frame(results_cor,stringsAsFactors = F)
results_cor<-results_cor[order(results_cor$signature,results_cor$q.value),]
results_cor_filt<-results_cor[results_cor$q.value<0.05,]
print("Number of significant hits for CRISPR screen:")
table(results_cor_filt$signature)
length(unique(results_cor_filt$gene))
```

# CRISPR correlated with drug response of target matched compound
```{r}
KO_drug_response<-corDrugResponse(results_cor_filt,-0.1)
KO_drug_response_qval<-KO_drug_response[KO_drug_response$q.value<0.01,] #38 sign hits
table(KO_drug_response_qval$signature)
```

#Correlation between RNAi gene essentiality score and signature activity
```{r warning=F, cache=T}
results_kd<-c()
for(j in 1:ncol(cell_exp_filt_KD))
{
  res_sig<-c()
  for(i in 1:ncol(gene_knockdown_filt))
  {
    res<-cor.test(as.numeric(gene_knockdown_filt[,i]),cell_exp_filt_KD[,j],method="k",alternative = "less")
    if(res$estimate< -0.07)
    {
      res_sig<-rbind(res_sig,c(j,colnames(gene_knockdown_filt)[i],
                                       res$estimate,res$p.value))
    }
  }
  res_sig<-data.frame(res_sig,stringsAsFactors = F)
  res_sig<-cbind(res_sig,q.value=p.adjust(res_sig[,4],method="BH"))
  results_kd<-rbind(results_kd,res_sig)
}

colnames(results_kd)<-c("signature","gene","rho","p.value","q.value")
results_kd<-data.frame(results_kd,stringsAsFactors = F)
results_kd<-results_kd[order(results_kd$signature,results_kd$q.value),]
results_kd_filt<-results_kd[results_kd$q.value<0.05,]
print("Number of significant hits for RNAi screen:")
table(results_kd_filt$signature)
length(unique(results_kd_filt$gene))
```

# RNAi correlated with drug response of target matched compound
```{r}
KD_drug_response<-corDrugResponse(results_kd_filt,-0.1)
KD_drug_response_qval<-KD_drug_response[KD_drug_response$q.value<0.01,] #76 sign hits
table(KD_drug_response_qval$signature)
```

##Combined results
```{r}
comb_res<-inner_join(results_cor_filt,results_kd_filt,by=c("signature","gene"))
print("Number of hits with support from both CRISPR and RNAi screen:")
table(comb_res$signature)
```

#Filter combined hits for significant overlaps
```{r}
total_genes<-sum(colnames(gene_knockdown_filt)%in%colnames(gene_effect_filt))
res<-c()
for(s in unique(comb_res$signature))
{
  KD_hits<-sum(results_kd_filt[results_kd_filt$signature==s,"gene"]%in%colnames(gene_effect_filt))
  KO_hits<-sum(results_cor_filt[results_cor_filt$signature==s,"gene"]%in%colnames(gene_knockdown_filt))
  overlap<-nrow(comb_res[comb_res$signature==s,])
  p<-phyper(overlap-1,KD_hits,total_genes-KD_hits,KO_hits,lower.tail = FALSE)
  res<-rbind(res,c(s,p))
}
res<-data.frame(as.matrix(res),stringsAsFactors = F)
colnames(res)<-c("signature","pvalue")
comb_res_filt<-comb_res[comb_res$signature%in%res[res$pvalue<0.05,"signature"],]
table(comb_res_filt$signature)
```

#Combined CRISPR and RNAi correlated with drug response of target matched compound
```{r}
comb_drug_response<-corDrugResponse(comb_res_filt,-0.07)
comb_drug_response_qval<-comb_drug_response[comb_drug_response$q.value<0.1,]
```

#Combined CRISPR and RNAi not correlated with drug response of target matched compound
```{r}
noresponse<-c()
for(i in 1:nrow(comb_res_filt))
{
  gname<-strsplit(comb_res_filt$gene[i],"\\.")[[1]][1]
  drugs<-drug_lookup[drug_lookup$target==gname,"drug"]
  sig<-comb_res_filt$signature[i]
  for(d in drugs)
  {
    pdat<-left_join(drug_screen[drug_screen$name==d,],
                    data.frame(depmap_id=rownames(cell_exp_filt_drug),cell_exp_filt_drug),
                    by="depmap_id")
    pdat<-left_join(pdat,cell_line_info,by="depmap_id")
    pdat<-pdat[pdat$passed_str_profiling.y==T,]
    pdat<-pdat[!is.na(pdat[,paste0("CX",sig)]),]
    res<-cor.test(pdat[,"auc"],pdat[,paste0("CX",sig)],method="k",alternative = "less")
    if(res$estimate >= -0.07)
    {
      noresponse<-rbind(noresponse,c(sig,gname,d,res$estimate,res$p.value))
    }
  }
}

colnames(noresponse)<-c("signature","target","drug","tau","p.value")
noresponse<-data.frame(noresponse,stringsAsFactors = F)
comb_drug_noresponse<-noresponse[order(noresponse$signature),] #209
table(comb_drug_noresponse$signature)
```

#Save results of the analyses
```{r}
write.table(KO_drug_response_qval, file=paste0(BASE,"/data/KO_drug_response_qval.txt"), sep="\t", row.names=F)
write.table(KD_drug_response_qval, file=paste0(BASE,"/data/KD_drug_response_qval.txt"), sep="\t", row.names=F)
write.table(comb_drug_response_qval, file=paste0(BASE,"/data/comb_drug_response_qval.txt"), sep="\t", row.names=F)
```

#Reduce list of biomarkers with non-significant drug
```{r}
target<-unique(drug_screen[,c("target")])
target<-unique(unlist(strsplit(target,", ")))
genes<-unique(sapply(strsplit(comb_res_filt$gene,"\\."),"[[",1))

nodrug_genes <- as.data.frame(genes[!genes %in% target]) #279
colnames(nodrug_genes)[1]<-"genes"
write.table(nodrug_genes, file=paste0(BASE,"/data/genes_nodrug.txt"), sep="\t", row.names=F, quote=F)
```